---
title: "DMP Figure"
author: "Andrés Delgado Galán"
date: "2026-03-16"
output: html_document
---

```{r}
library(here)

# avena_CpG_GUvsUU <- readRDS(here("avena_ncbi","DMPs_CpG_GUvsUU.rds"))
# avena_CpG_UUvsUC <- readRDS(here("avena_ncbi","DMPs_CpG_UUvsUC.rds"))
# avena_CpG_GUvsGC <- readRDS(here("avena_ncbi","DMPs_CpG_GUvsGC.rds"))
# 
# avena_CHG_GUvsUU <- readRDS(here("avena_ncbi","DMPs_CHG_GUvsUU.rds"))
# avena_CHG_UUvsUC <- readRDS(here("avena_ncbi","DMPs_CHG_UUvsUC.rds"))
# avena_CHG_GUvsGC <- readRDS(here("avena_ncbi","DMPs_CHG_GUvsGC.rds"))
# 
# avena_CHH_GUvsUU <- readRDS(here("avena_ncbi","DMPs_CHH_GUvsUU.rds"))
# avena_CHH_UUvsUC <- readRDS(here("avena_ncbi","DMPs_CHH_UUvsUC.rds"))
# avena_CHH_GUvsGC <- readRDS(here("avena_ncbi","DMPs_CHH_GUvsGC.rds"))
# 
# plantago_CpG_GUvsUU <- readRDS(here("plantago_pseudo","DMPs_CpG_GUvsUU.rds"))
# plantago_CpG_UUvsUC <- readRDS(here("plantago_pseudo","DMPs_CpG_UUvsUC.rds"))
# plantago_CpG_GUvsGC <- readRDS(here("plantago_pseudo","DMPs_CpG_GUvsGC.rds"))
# 
# plantago_CHG_GUvsUU <- readRDS(here("plantago_pseudo","DMPs_CHG_GUvsUU.rds"))
# plantago_CHG_UUvsUC <- readRDS(here("plantago_pseudo","DMPs_CHG_UUvsUC.rds"))
# plantago_CHG_GUvsGC <- readRDS(here("plantago_pseudo","DMPs_CHG_GUvsGC.rds"))
# 
# plantago_CHH_GUvsUU <- readRDS(here("plantago_pseudo","DMPs_CHH_GUvsUU.rds"))
# plantago_CHH_UUvsUC <- readRDS(here("plantago_pseudo","DMPs_CHH_UUvsUC.rds"))
# plantago_CHH_GUvsGC <- readRDS(here("plantago_pseudo","DMPs_CHH_GUvsGC.rds"))

CpG_meth <- readRDS(here("avena_ncbi","CpG_meth.rds"))
CHG_meth <- readRDS(here("avena_ncbi","CHG_meth.rds"))
CHH_meth <- readRDS(here("avena_ncbi","CHH_meth.rds"))
```


```{r CpG DIFFERENTIAL METHYLATION ANALYSIS, include=FALSE}
# I will create a function to then compare all against all

# Function for pairwise comparison
meth_diff <- 10
p_value <- 0.05
pairwise_diff <- function(meth_obj, g1, g2, min_diff = meth_diff, q_cutoff = p_value) {
  
  # Subset the object to only keep the two groups of interest
  sub_obj <- reorganize(
    meth_obj,
    sample.ids = getSampleID(meth_obj)[meth_obj@treatment %in% c(g1, g2)],
    treatment = meth_obj@treatment[meth_obj@treatment %in% c(g1, g2)]
  )
  
  # Run differential methylation
  diff <- calculateDiffMeth(
    sub_obj,
    overdispersion = "MN",
    test = "Chisq"
  )
  
  # Extract significant CpGs
  sig <- getMethylDiff(diff, difference = min_diff, qvalue = q_cutoff)
  
  return(list(all = diff, significant = sig))
}


# MAKE COMPARISONS

# GRAZED VS. UNGRAZED (WITHIN UNCLIPPED)
res_GUvsUU <- pairwise_diff(CpG_meth, g1 = 2, g2 = 3)
# UNCLIPPED VS. CLIPPED (WITHIN GRAZED)
res_GUvsGC <- pairwise_diff(CpG_meth, g1 = 2, g2 = 0)
 # UNCLIPPED VS. CLIPPED (WITHIN UNGRAZED)
 res_UUvsUC <- pairwise_diff(CpG_meth, g1 = 3, g2 = 1)
```


## Differential methylation analysis



Con la función "pairwise_diff()" de methylKit, podemos comparar el estado de metilación de cada citosina entre la población de cada tratamiento y la población control (Grazed es el control de Ungrazed y Unclipped es el control de Clipped). Obtenemos hipo- e hiper- metilaciones con respecto al control, y cada una de estas diferencias está asociada a un p-valor. En este caso hemos considerado como Differentially Methylated Positions (**DMPs**) cuando la **diferencia de metilación \> `r meth_diff`%** y el **p-valor \< `r p_value`**


```{r CpG exporting table of DMPs, include=FALSE}

# Apply function to all your result objects
DMPs_CpG_GUvsUU <- res_GUvsUU
DMPs_CpG_GUvsGC <- res_GUvsGC
DMPs_CpG_UUvsUC <- res_UUvsUC

# Save objects
saveRDS(DMPs_CpG_GUvsUU, "avena_DMPs_CpG_GUvsUU.rds")
saveRDS(DMPs_CpG_GUvsGC, "avena_DMPs_CpG_GUvsGC.rds")
saveRDS(DMPs_CpG_UUvsUC, "avena_DMPs_CpG_UUvsUC.rds")
```


```{r CpG VOLCANO PLOTS FOR VISUALIZING DMPs, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Volcano plot function
plot_volcano <- function(diff_obj, min_diff = 10, q_cutoff = 0.05, title = "") {
  df <- as.data.frame(diff_obj)
  
  # Calculate variables
  df$logq <- -log10(df$qvalue)
  df$status <- ifelse(abs(df$meth.diff) >= min_diff & df$qvalue <= q_cutoff, "sig", "ns")
  
  # Plot
  p <- ggplot(df, aes(x = meth.diff, y = logq)) +
    geom_point(aes(color = status), alpha = 0.6, size = 1.5, show.legend = FALSE) +
    scale_color_manual(values = c("sig" = "red", "ns" = "black")) +
    geom_vline(xintercept = c(-min_diff, min_diff), linetype = "dashed", color = "blue") +
    geom_hline(yintercept = -log10(q_cutoff), linetype = "dashed", color = "blue") +
    labs(
      x = "% Methylation Difference",
      y = expression(-log[10]("q-value")),
      title = title
    ) +
    theme_bw() +
    theme(
      axis.title = element_blank(),  # remove axis labels
      axis.text = element_text(size = 9),
      plot.title = element_text(size = 11, hjust = 0.5)
    )
  
  return(p)
}

# Generate volcano plots
p1 <- plot_volcano(res_GUvsUU$all, title = "Ungrazed (Unclipped)")
p2 <- plot_volcano(res_GUvsGC$all, title = "Clipping (Grazed)")
p3 <- plot_volcano(res_UUvsUC$all, title = "Clipping (Ungrazed)")


grid.arrange(
  p1, p2, p3, ncol = 3,
  left   = textGrob(expression(-log[10](q-value)), rot = 90, gp = gpar(fontsize = 12)),
  bottom = textGrob("% Methylation Difference", gp = gpar(fontsize = 12))
)

```

```{r CpG DMPs table, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Extract the significant results
sigUngrazed <- res_GUvsUU$significant
sigClipped <- res_GUvsGC$significant
sigUngrazedClipped <- res_UUvsUC$significant

# Count positives and negatives
n_posUngrazed <- sum(sigUngrazed$meth.diff > 0)
n_negUngrazed <- sum(sigUngrazed$meth.diff < 0)

n_posClipped <- sum(sigClipped$meth.diff > 0)
n_negClipped <- sum(sigClipped$meth.diff < 0)

n_posUngrazedClipped <- sum(sigUngrazedClipped$meth.diff > 0)
n_negUngrazedClipped <- sum(sigUngrazedClipped$meth.diff < 0)


# Build the table
meth_table <- data.frame(
  Condition = c("Hypermethylated", "Hypomethylated"),
  UngrazedU = c(n_posUngrazed, n_negUngrazed),
  ClippingG = c(n_posClipped, n_negClipped),
  ClippingU = c(n_posUngrazedClipped, n_negUngrazedClipped)
)

# Show nice table in R Markdown
knitr::kable(meth_table, align = "c", caption = "Differentially methylated sites")

```

