---
title: "A.barbata CHH NCBI"
author: "Andrés Delgado"
output: html_document
---

En este documento presento algunos resultados sobre los patrones de metilación de Avena bajo distintos tratamientos de herbivoría. He usado el mismo flujo de trabajo que en avena_pseudo_results.Rmd con la diferencia de que ahora trabajamos con un genoma de referencia completo, en lugar de usar el pseudogenoma que preparé yo con stacks.


## PREPROCESSING

#### Brevemente resumo los pasos previos al análisis:

1.  Se utilizó como genoma de referencia el ensamblaje Abarbata_PI388828_v1.0 (Avena barbata, GCA_947310925.1), descargado de NCBI Genome el [02/09/2025] en el siguiente enlace: https://www.ncbi.nlm.nih.gov/datasets/genome/GCA_947310925.1/

2.  Alineamiento de las lecturas de bisulfito de mis muestras tratadas con el genoma de referencia usando BISMARK en linux.

3.  Generación de informes de metilación de las muestras asociados a sus posiciones en el genoma de referencia. BISMARK genera estos informes en forma de archivos .bam

4.  Importación de los archivos .bam a methylkit para el análisis


```{r setup LOADING DATA AND PACKAGES, include=FALSE}
#Here I paste the path to the original files which is not inside the project due to their great size
# bam_dir <- "C:/Users/iglorena.IRNASE/Documents/andres/INTERTROPHIC/EPIGENETICA/NOVOGENE_DOWNLOAD/methylation_report_avena/ncbi_genome_pe"
# 
#  bam.list <- as.list(list.files(
#    path   = bam_dir,
#    pattern = "_bismark_bt2_pe.sorted\\.bam$",
#    full.names = FALSE
#  ))
# 
#  # Save list as an R object
#  saveRDS(bam.list, "bam_list.rds")

# Read bam.list
bam.list <- readRDS("bam_list.rds")


# --- Core packages ---
library(here)        # Reproducibilidad de rutas
library(methylKit)   # Análisis de metilación
library(tidyverse)   # Incluye ggplot2, dplyr, tidyr, readr, purrr, stringr, forcats, tibble

# --- Visualización y clustering ---
library(pheatmap)    # Mapas de calor y correlaciones
library(dendextend)  # Manipulación y mejora de dendrogramas
library(gridExtra)   # Combinar varios plots (más potente que grid)
library(grid)        # Gráficos base de R (necesario a veces para layout)
library(kableExtra)  # Tablas bonitas en RMarkdown

# --- Genética ---
library(adegenet)    # DAPC y análisis genético
library(vcfR)        # read and manipulate VCF files
library(vegan)       # Procrustes analysis
```

```{r EXTRACTING COMMON INFORMATION FOR ALL CONTEXT, include=FALSE}
# sample.id assigns a unique ID to each sample (we remove the suffix and have the id)
sample.id <- as.list(gsub("_R1_trimmed_bismark_bt2_pe.sorted\\.bam$", "", bam.list))

# treatment vector specifies experimental group
# Usually methylKit is designed for binary treatment designs
# so i will need to encode both factors into a single one
treatment_table <- read.table(file = here("avena_treatment.csv"), header = T, sep = ";")


# # CHECKING THAT TREATMENT TABLE AND SAMPLE LIST ARE THE SAME
# GET THE VECTORS
sample_id_vec <- unlist(sample.id)
table_ids <- treatment_table$id

 # Find ids in sample.id that are NOT in treatment_table$id
 ids_in_sample_not_in_table <- setdiff(sample_id_vec, table_ids)

 # Find ids in treatment_table$id that are NOT in sample.id
 ids_in_table_not_in_sample <- setdiff(table_ids, sample_id_vec)

 # Print results
 cat("IDs in sample.id but not in treatment_table$id:\n")
 print(ids_in_sample_not_in_table)

 cat("IDs in treatment_table$id but not in sample.id:\n")
 print(ids_in_table_not_in_sample)
 
 # Remove row where id is "AV275"
treatment_table <- treatment_table %>%
  filter(!id %in% c(
    "AV253_2", "AV243_2", "AV282_2",
    "AV280_2", "AV297_2", "AV275", "AV296_2"
  ))


# Reorder treatment_table so its ids follow the order in sample.id
treatment_table <- treatment_table[match(sample_id_vec, treatment_table$id), ]

# Convert columns to factors
treatment_table$maternal <- factor(treatment_table$maternal, levels = c("graz", "ungraz"))
treatment_table$simulation <- factor(treatment_table$simulation, levels = c("clip", "unclip"))

# Combine into a single factor
treatment_table$combined <- interaction(treatment_table$maternal, treatment_table$simulation, sep = "-")
levels(treatment_table$combined)
# [1] "graz-clip" "ungraz-clip" "graz-unclip" "ungraz-unclip"

# encode this as numeric for methylKit (e.g., 0, 1, 2, 3)
treatment_multi <- as.numeric(treatment_table$combined) - 1
# Now treatment has values: 0, 1, 2, 3 (corresponding to the 4 groups)
```

```{r CHH CREATING and SAVING METHYLKIT OBJECT, include=FALSE}
#THIS STEP TAKES TOO LONG SO I WILL ONLY RUN IT WHEN I NEED TO CHANGE IT
# # full paths → required by processBismarkAln
# bam.path <- file.path(bam_dir, bam.list)
# bam.path  <- as.character(bam.path)
# sample.id <- as.character(sample.id)
# 
# 
# # process each BAM individually
# # We use this loop because we can't change directory in Rmarkdown, and processBismarkAln doesn't accept a vector of bam files, it needs them one at a time
# objs <- mapply(
#   function(file, id, tr) {
#     processBismarkAln(
#       location   = file,
#       sample.id  = id,
#       assembly   = "mock",
#       treatment  = tr,
#       mincov = 0,
#       read.context = "CHH"
#     )
#   },
#   file = bam.path,
#   id   = sample.id,
#   tr   = treatment_multi,
#   SIMPLIFY = FALSE
# )
# 
# # merge them into one methylRawList
# avena_CHH_obj <- new("methylRawList", objs)
# 
# # set treatment vector explicitly
# # even though we already specified when creating the object, using the loop didn't work for treatment
# avena_CHH_obj@treatment <- treatment_multi
# 
#  # Save the result for later use
#  saveRDS(avena_CHH_obj, file = "avena_ncbi_CHH.rds")
```

```{r CHH LOADING, FILTERING, UNITING, include=FALSE}

# --- 1. Cargar y unir ---
avena_CHH_obj <- readRDS(here("avena_ncbi", "avena_ncbi_CHH.rds"))


# --- 1. Unificar muestras ---
CHH_meth_pre_filter <- methylKit::unite(avena_CHH_obj, min.per.group = 0L)
meth_pre <- getData(CHH_meth_pre_filter)

# --- 1b. Filtro de cobertura máxima ---
# Definir umbrales de cobertura
minimum_coverage <- 5
maximum_coverage <- 1000


# Identificar columnas de cobertura
cov_cols <- grep("^coverage", names(meth_pre))

# Calcular cuántos sitios se quedan fuera de los filtros de cobertura
n_sites_under_cov <- sum(apply(meth_pre[, cov_cols], 1, function(x) any(x < minimum_coverage, na.rm = TRUE)))
n_sites_over_cov <- sum(apply(meth_pre[, cov_cols], 1, function(x) any(x > maximum_coverage, na.rm = TRUE)))


# Mantener solo sitios con cobertura entre minimum_coverage y maximum_coverage en todas las muestras
CHH_meth_coverage <- CHH_meth_pre_filter[
  rowMeans(
    getData(CHH_meth_pre_filter)[, cov_cols] > maximum_coverage |
    getData(CHH_meth_pre_filter)[, cov_cols] < minimum_coverage,
    na.rm = TRUE
  ) == 0,
]


# --- 2. Filtrado de sitios polimórficos ---
meth_cov <- getData(CHH_meth_coverage)

numC_cols <- grep("^numC", names(meth_cov))
numT_cols <- grep("^numT", names(meth_cov))

meth_cov <- meth_cov %>%
  mutate(
    has_C = rowSums(across(all_of(numC_cols), ~ .x > 0), na.rm = TRUE) > 0,
    has_T = rowSums(across(all_of(numT_cols), ~ .x > 0), na.rm = TRUE) > 0,
    polymorphic = has_C & has_T
  )

CHH_meth_filtered_polymorphic <- CHH_meth_coverage[meth_cov$polymorphic, ]

# --- 3. Filtrado por MEF (Minor Epiallele Frequency, a nivel poblacional) ---
mef_threshold <- 0.01
meth <- getData(CHH_meth_filtered_polymorphic)

# --- Seleccionar columnas numC y numT ---
numC_cols <- grep("^numC", names(meth), value = TRUE)
numT_cols <- grep("^numT", names(meth), value = TRUE)

numC_mat <- as.matrix(meth[, numC_cols])
numT_mat <- as.matrix(meth[, numT_cols])

# --- Calcular sumas globales (por CHH) ---
sum_C <- rowSums(numC_mat, na.rm = TRUE)
sum_T <- rowSums(numT_mat, na.rm = TRUE)

# --- Minoritario global y total global ---
minor_global <- pmin(sum_C, sum_T)
total_global <- sum_C + sum_T

# --- Calcular MEF poblacional ---
mef_pop <- minor_global / total_global

# --- Crear vector lógico con el umbral ---
keep_mef <- mef_pop >= mef_threshold
keep_mef[is.na(keep_mef)] <- FALSE

# --- Filtrar objeto original ---
CHH_meth_filtered_mef <- CHH_meth_filtered_polymorphic[keep_mef, ]

# --- Reporte resumen ---
message("Sitios totales antes del filtrado: ", nrow(CHH_meth_filtered_polymorphic))
message("Sitios retenidos tras el filtrado MEF >= ", mef_threshold, ": ", sum(keep_mef))
message("Porcentaje retenido: ", round(100 * mean(keep_mef), 2), "%")


# --- 4. Filtrado por missingness dentro de cada grupo ---
meth <- getData(CHH_meth_filtered_mef)
treatment <- CHH_meth_filtered_mef@treatment
cov_cols <- grep("^coverage", names(meth))

# Agrupar índices de columnas por tratamiento
cov_groups <- split(cov_cols, treatment)

# Calcular proporción de sitios con datos (no NA) por grupo
group_prop_df <- map_dfc(cov_groups, function(cols) {
  tibble(prop = rowMeans(!is.na(meth[, cols, drop = FALSE])))
}) %>%
  set_names(paste0("group_", names(cov_groups)))

# --- 5. Aplicar filtro de cobertura mínima ---
threshold <- 0.8
keep <- group_prop_df %>%
  mutate(pass = apply(., 1, function(x) all(x >= threshold))) %>%
  pull(pass)

CHH_meth_80_by_group <- CHH_meth_filtered_mef[keep, ]
CHH_meth <- CHH_meth_80_by_group

# Objeto final
meth <- getData(CHH_meth)

# IDs de las muestras
sample_ids <- CHH_meth@sample.ids

# columnas de interés para sacar el porcentaje de NAs en cada muestra
cov_cols <- grep("^coverage", names(meth))

# porcentaje de NAs en cada muestra
na_percent <- sapply(cov_cols, function(col) {
  mean(is.na(meth[, col])) * 100  # porcentaje de filas con NA
})

# Asociarlo a los sample IDs
na_percent_df <- data.frame(
  sample_id = sample_ids,
  pct_missing = na_percent
)


```

## SMP calling and filtering 

We used "processBismarkAln" function of the methylKit R package to call all the cytosines in CHH position reported in the bam files. We removed `r n_sites_under_cov` positions with coverage under `r minimum_coverage` and `r n_sites_over_cov` positions with coverage above `r maximum_coverage`. This filtering step eliminated CHH sites with insufficient coverage to provide reliable methylation estimates, as well as sites with excessively high coverage likely arising from artifacts such as collapsed repetitive regions. From a total of **`r nrow(CHH_meth_coverage) `** positions, **`r nrow(CHH_meth_filtered_polymorphic) `** were polymorphic and therefore included in our SMP dataset. Additionally, SMPs with Minor Epiallele Frequency (MEF) ≤ `r mef_threshold * 100`% were removed as rare, uninformative, and potentially unreliable. Finally, we filtered the resulting **`r nrow(CHH_meth_filtered_mef) `** SMPs to keep only those present in at least `r threshold * 100`% of the individuals of every treatment (i.e., combinations of maternal environment and herbivory simulation treatments). After all the filters, we checked the missing data for every sample and the maximum was `r round(max(na_percent_df$pct_missing), 1)`%. Thus, we considered unnecessary to filter by sample, resulting in **`r nrow(CHH_meth_80_by_group) `** SMPs ready to be analysed.

```{r CHH Plotting coverage, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Extraer los datos de cobertura
meth_pre <- getData(CHH_meth_pre_filter)
cov_cols <- grep("^coverage", names(meth_pre), value = TRUE)

# Pasar a formato largo para ggplot
coverage_long <- meth_pre[, cov_cols] %>%
  reshape2::melt(variable.name = "sample", value.name = "coverage") %>%
  filter(!is.na(coverage), coverage > 0)

# Contar SMPs por cobertura real
coverage_counts <- coverage_long %>%
  group_by(sample, coverage) %>%
  summarise(n = n(), .groups = "drop")

# 2. Graficar
ggplot(coverage_counts, aes(x = coverage, y = n, color = sample)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 0.7) +
  geom_vline(xintercept = minimum_coverage, color = "black", linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = maximum_coverage, color = "black", linetype = "dashed", linewidth = 0.8) +
  scale_x_log10(
    breaks = c(1, 2, 3, 4, 5, 10, 20, 50, 100, 500, 1000, 5000, 10000, 100000),
    labels = scales::comma_format(accuracy = 1)
  ) +
  scale_y_log10(
    breaks = c(0, 1, 2, 5,10, 50, 100, 500, 1000, 5000, 10000, 50000, 100000, 500000),
    labels = scales::comma_format(accuracy = 1)
  ) +
  labs(
    x = "Coverage",
    y = "SMP count",
    title = "Coverage filter"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

```

```{r CHH Plotting epiallele frequency, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
mef_df <- data.frame(MEF_pop = mef_pop) %>%
  filter(!is.na(MEF_pop)) %>%
  mutate(mef_percent = MEF_pop * 100)

h <- hist(mef_df$mef_percent,
     breaks = 50,  # adjust number of bins
     main = "MEF filter",
     xlab = "Minor Epiallele Frequency (%)",
     ylab = "SMP count",
     col = "skyblue",
     border = "white")
     

# Add vertical line at x = 1
abline(v = 1, col = "red", lwd = 2, lty = 2)  # lwd = line width, lty = line type (dashed)

```


## Clustering methylation profiles

El primer paso en el análisis ha sido explorar visualmente las diferencias en los perfiles de metilación entre tratamientos. Para ello hemos utilizado los porcentajes de metilación de cada una de las posiciones CHH (posiciones del genoma con una Citosina seguida dos bases distintas de Guanina). Hemos evaluado un total de **`r nrow(CHH_meth)`** posiciones compartidas por las **`r (ncol(CHH_meth) - 4)/3 `** muestras, separando por tratamientos para ver posibles agrupaciones. Aquí muestro un Correlograma, un Dendograma y un DAPC (Discriminant Analysis of Principal Components), que es un análisis multivariante usado en genética de poblaciones para describir y visualizar la estructura genética entre grupos.Utiliza el PCA y luego maximiza la separación entre grupos predefinidos (los tratamientos).

```{r CHH METHYLATION % MATRIX, include=FALSE}
CHH_meth_matrix <- as.matrix(percMethylation(CHH_meth))
# extract genomic positions
coords <- getData(CHH_meth)$start

# assign them as rownames to meth_matrix
rownames(CHH_meth_matrix) <- coords
```

```{r CHH CORRELOGRAM, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Correlation matrix
corr_matrix <- cor(CHH_meth_matrix, use = "pairwise.complete.obs")

# Visualization
pheatmap(corr_matrix,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
         main = "Methylation profile Correlations",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation")
```

```{r CHH CLUSTERING, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Perform hierarchical clustering of samples based on methylation profiles using correlation distance and Ward's method, then plot

# Run clusterSamples without plotting
hc <- clusterSamples(CHH_meth, dist = "correlation", method = "ward.D2", plot = FALSE)


# Sample IDs
sample_ids <- CHH_meth@sample.ids   # correct names of your samples

# Treatments
treatments <- getTreatment(CHH_meth)  # same as CHH_meth@treatment

# Map colors
treatment_colors <- c("0" = "#FF69B4", "1" = "#00FF7F", "2" = "#8B3A62", "3" = "#008B45")

# Convert hclust to dendrogram
dend <- as.dendrogram(hc)

# Reorder colors according to dendrogram tip labels
tip_order <- labels(dend)  # these should be sample_ids if clusterSamples worked correctly
tip_colors <- treatment_colors[as.character(treatments[match(tip_order, sample_ids)])]

# Assign colors
labels_colors(dend) <- tip_colors

# Plot
plot(dend, main = "Methylation profile clustering")
```

```{r CHH PREPARING DAPC, echo=FALSE, fig.show='hide'}
# extract genomic positions
coords <- getData(CHH_meth)$start

# assign them as rownames to meth_matrix
rownames(CHH_meth_matrix) <- coords

grp <- as.factor(treatments)


# VAMOS A IMPUTAR VALORES PARA SUSTITUIR NAs QUE DAPC NO PUEDE MANEJAR
# USAMOS LA MEDIANA DE CADA FILA PARA RELLENAR LOS NAs DE ESA FILA
CHH_meth_matrix_imputed <- CHH_meth_matrix
for (i in 1:nrow(CHH_meth_matrix_imputed)) {
  CHH_meth_matrix_imputed[i, is.na(CHH_meth_matrix_imputed[i, ])] <- 
    median(CHH_meth_matrix_imputed[i, ], na.rm = TRUE)
}



# Run DAPC keeping 14 principal components (and all possible discriminant functions). I used 14 because it's a safe maximum: 42/3
dapc_res <- dapc(t(CHH_meth_matrix_imputed), grp, n.pca = 14, n.da = length(levels(grp)) - 1)

# Optimize the number of PCs using the alpha-score method (balances discrimination vs overfitting).
optim_a <- optim.a.score(dapc_res)

# Print the optimal number of PCs suggested by optim.a.score
cat("Optimal number of PCs (alpha-score):", optim_a$best, "\n")

# Re-run DAPC using the optimal number of PCs suggested automatically
dapc_res <- dapc(
  t(CHH_meth_matrix_imputed), 
  grp,
  n.pca = max(2, optim_a$best),
  n.da  = length(levels(grp)) - 1
)

# Extract full loadings matrix (all CHHs × LDs)
loadings <- dapc_res$var.contr  

# Convert to tidy data frame: CHH, LD, Loading
loadings_df <- reshape2::melt(loadings,
                    varnames = c("CHH", "LD"),
                    value.name = "Loading")

# Save to tab-delimited text file
write.table(loadings_df, "avena_ncbi_CHH_loadings_with_positions.csv",
            sep = ";", quote = FALSE, row.names = FALSE)

# ---- PCA instead of DAPC ----
pca_res <- prcomp(t(CHH_meth_matrix_imputed), scale. = TRUE)

```

```{r CHH PLOTTING DAPC, echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
# cREATING A DATAFRAME WITH DAPC, TREATMENT AND SAMPLE ID
dapc_scores <- as.data.frame(dapc_res$ind.coord)
dapc_scores$id <- sample_ids
dapc_scores$combined <- grp

# Correct factor levels and labels (matching your desired order and names)
dapc_scores$combined <- factor(dapc_scores$combined,
                               levels = c("0", "2", "1", "3"),
                               labels = c("Grazed + Clipped",
                                          "Grazed + Unclipped",
                                          "Ungrazed + Clipped",
                                          "Ungrazed + Unclipped"))

# Compute group centroids
centroids <- dapc_scores %>%
  group_by(combined) %>%
  summarise(LD1 = mean(LD1), LD2 = mean(LD2))

# Plot
ggplot(dapc_scores, aes(x = LD1, y = LD2, color = combined)) +
  # Vectors from centroid to each point
  geom_segment(data = dapc_scores %>%
                 left_join(centroids, by = "combined"),
               aes(x = LD1.y, y = LD2.y, xend = LD1.x, yend = LD2.x, color = combined),
               alpha = 0.4, size = 0.6, inherit.aes = FALSE) +
  # Points
  geom_point(size = 1.8) +
  # Open inertia/confidence ellipses
  stat_ellipse(type = "norm", level = 0.67, size = 0.6) +
  # Colors for each group
  scale_color_manual(values = c(
    "Grazed + Clipped" = "#FF69B4",
    "Grazed + Unclipped" = "#8B3A62",
    "Ungrazed + Clipped" = "#00FF7F",
    "Ungrazed + Unclipped" = "#008B45"
  )) +
  # Legend title
  labs(title = "DAPC of CHH Methylation in A.barbata",
       x = "LD1", y = "LD2",
       color = "Herbivory treatment") +
  # Theme adjustments
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),   # remove grid
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1), # box around plot
    aspect.ratio = 1  # make the plot square
  )

```

```{r CHH PLOTTING PCA, echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
# Create dataframe with PCA scores
pca_scores <- as.data.frame(pca_res$x)
pca_scores$id <- sample_ids
pca_scores$combined <- grp

# Correct factor levels and labels
pca_scores$combined <- factor(pca_scores$combined,
                               levels = c("0", "2", "1", "3"),
                               labels = c("Grazed + Clipped",
                                          "Grazed + Unclipped",
                                          "Ungrazed + Clipped",
                                          "Ungrazed + Unclipped"))

# ---- Compute group centroids ----
centroids <- pca_scores %>%
  group_by(combined) %>%
  summarise(PC1 = mean(PC1), PC2 = mean(PC2))

# ---- Plot PCA with centroids and segments ----
ggplot(pca_scores, aes(x = PC1, y = PC2, color = combined)) +
  
  # Segments from centroid → each point
  geom_segment(data = pca_scores %>%
                 left_join(centroids, by = "combined"),
               aes(x = PC1.y, y = PC2.y,
                   xend = PC1.x, yend = PC2.x,
                   color = combined),
               alpha = 0.4, size = 0.6, inherit.aes = FALSE) +

  # Points
  geom_point(size = 1.8) +
  
  # Ellipses
  stat_ellipse(type = "norm", level = 0.67, size = 0.6) +
  
  # Colors
  scale_color_manual(values = c(
    "Grazed + Clipped" = "#FF69B4",
    "Grazed + Unclipped" = "#8B3A62",
    "Ungrazed + Clipped" = "#00FF7F",
    "Ungrazed + Unclipped" = "#008B45"
  )) +

  # Labels with explained variance
  labs(title = "PCA of CHH Methylation in P. lagopus",
       x = paste0("PC1 (", round(summary(pca_res)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_res)$importance[2,2]*100, 1), "%)"),
       color = "Herbivory treatment") +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    aspect.ratio = 1
  )
```

## SNP to SMP correlation
Patterns of epigenetic variation (SMPs) among individuals and populations could be similar to those of genetic variation (SNPs). To evaluate this we plot a similar DAPC with SNP data and look at the correlation between both DAPCs using procrustes analysis.

```{r CHH PREPARING SNP DAPC, include=FALSE}
# read VCF
vcf <- read.vcfR(here("avena_ncbi", "filtered_variants_cov5-1000_biallelic_MAF0.01_80pct.vcf.gz"))

# Convert to genlight (handles large SNP sets efficiently)
gl <- vcfR2genlight(vcf)    # converts to genlight; ploidy defaults to 2
indNames(gl) <- sub("_R1_trimmed_bismark_bt2_pe\\.sorted$", "", indNames(gl))

# set population/factors for samples
pop(gl) <- treatments

# 3) missing data handling / imputation (replace NA by locus median)
# compute allele dosage matrix
dos <- as.matrix(gl)  # genlight -> numeric matrix (0/1/2), NAs remain
# Remove some monomorphic positions
keep <- apply(dos, 2, function(x) {
    u <- unique(na.omit(x))
    length(u) > 1
})
sum(!keep)    # should show 79
dos_filtered <- dos[, keep]

# impute by median allele dosage per SNP:
for(j in seq_len(ncol(dos_filtered))){
  nas <- is.na(dos_filtered[,j])
  if(any(nas)) dos_filtered[nas,j] <- median(dos_filtered[,j], na.rm=TRUE)
}



# put back into a genlight
gl_imputed <- new("genlight", dos_filtered)
ploidy(gl_imputed) <- 2
pop(gl_imputed) <- pop(gl)

# Run DAPC keeping 14 principal components (and all possible discriminant functions). I used 14 because it's a safe maximum: 42/3
dapc_snp <- dapc(gl_imputed, grp, n.pca = 14, n.da = length(levels(grp)) - 1)

# Optimize the number of PCs using the alpha-score method (balances discrimination vs overfitting).
optim_a <- optim.a.score(dapc_snp)

# Print the optimal number of PCs suggested by optim.a.score
cat("Optimal number of PCs (alpha-score):", optim_a$best, "\n")

# Re-run DAPC using the optimal number of PCs suggested automatically
dapc_snp <- dapc(
  gl_imputed, 
  grp,
  n.pca = max(2, optim_a$best),
  n.da  = length(levels(grp)) - 1
)

# ---- PCA instead of DAPC ----
pca_snp_res <- prcomp(dos_filtered, scale. = TRUE)


```


```{r CHH PLOTTING SNP DAPC, echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
# cREATING A DATAFRAME WITH DAPC, TREATMENT AND SAMPLE ID
dapc_snp_scores <- as.data.frame(dapc_snp$ind.coord)
dapc_snp_scores$id <- sample_ids
dapc_snp_scores$combined <- grp

# Correct factor levels and labels (matching your desired order and names)
dapc_snp_scores$combined <- factor(dapc_snp_scores$combined,
                               levels = c("0", "2", "1", "3"),
                               labels = c("Grazed + Clipped",
                                          "Grazed + Unclipped",
                                          "Ungrazed + Clipped",
                                          "Ungrazed + Unclipped"))

# Compute group centroids
centroids <- dapc_snp_scores %>%
  group_by(combined) %>%
  summarise(LD1 = mean(LD1), LD2 = mean(LD2))

# Plot
ggplot(dapc_snp_scores, aes(x = LD1, y = LD2, color = combined)) +
  # Vectors from centroid to each point
  geom_segment(data = dapc_snp_scores %>%
                 left_join(centroids, by = "combined"),
               aes(x = LD1.y, y = LD2.y, xend = LD1.x, yend = LD2.x, color = combined),
               alpha = 0.4, size = 0.6, inherit.aes = FALSE) +
  # Points
  geom_point(size = 1.8) +
  # Open inertia/confidence ellipses
  stat_ellipse(type = "norm", level = 0.67, size = 0.6) +
  # Colors for each group
  scale_color_manual(values = c(
    "Grazed + Clipped" = "#FF69B4",
    "Grazed + Unclipped" = "#8B3A62",
    "Ungrazed + Clipped" = "#00FF7F",
    "Ungrazed + Unclipped" = "#008B45"
  )) +
  # Legend title
  labs(title = "DAPC of SNPs in A.barbata",
       x = "LD1", y = "LD2",
       color = "Herbivory treatment") +
  # Theme adjustments
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),   # remove grid
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1), # box around plot
    aspect.ratio = 1  # make the plot square
  )
```

```{r CHH PLOTTING SNP PCA, echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
# Create dataframe with PCA scores
pca_snp_scores <- as.data.frame(pca_snp_res$x)
pca_snp_scores$id <- sample_ids
pca_snp_scores$combined <- grp

# Correct factor levels and labels
pca_snp_scores$combined <- factor(pca_snp_scores$combined,
                               levels = c("0", "2", "1", "3"),
                               labels = c("Grazed + Clipped",
                                          "Grazed + Unclipped",
                                          "Ungrazed + Clipped",
                                          "Ungrazed + Unclipped"))

# ---- Compute group centroids ----
centroids <- pca_snp_scores %>%
  group_by(combined) %>%
  summarise(PC1 = mean(PC1), PC2 = mean(PC2))

# ---- Plot PCA with centroids and segments ----
ggplot(pca_snp_scores, aes(x = PC1, y = PC2, color = combined)) +
  
  # Segments from centroid → each point
  geom_segment(data = pca_snp_scores %>%
                 left_join(centroids, by = "combined"),
               aes(x = PC1.y, y = PC2.y,
                   xend = PC1.x, yend = PC2.x,
                   color = combined),
               alpha = 0.4, size = 0.6, inherit.aes = FALSE) +

  # Points
  geom_point(size = 1.8) +
  
  # Ellipses
  stat_ellipse(type = "norm", level = 0.67, size = 0.6) +
  
  # Colors
  scale_color_manual(values = c(
    "Grazed + Clipped" = "#FF69B4",
    "Grazed + Unclipped" = "#8B3A62",
    "Ungrazed + Clipped" = "#00FF7F",
    "Ungrazed + Unclipped" = "#008B45"
  )) +

  # Labels with explained variance
  labs(title = "PCA of SNPs in P. lagopus",
       x = paste0("PC1 (", round(summary(pca_snp_res)$importance[2,1]*100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_snp_res)$importance[2,2]*100, 1), "%)"),
       color = "Herbivory treatment") +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    aspect.ratio = 1
  )
```

```{r CHH PROCRUSTES ANALYSIS, include=FALSE}
# Select only numeric LD axes
dapc_axes <- dapc_scores[, grep("^LD", colnames(dapc_scores))]
dapc_snp_axes <- dapc_snp_scores[, grep("^LD", colnames(dapc_snp_scores))]

# Make sure row order is identical
all(rownames(dapc_axes) == rownames(dapc_snp_axes))  # should be TRUE

# Run Procrustes
proc <- procrustes(dapc_axes, dapc_snp_axes, scale = TRUE)

# Summary
summary(proc)

# Statistical test
proc_test <- protest(dapc_axes, dapc_snp_axes, permutations = 999)
proc_test
```

```{r CHH PROCRUSTES PLOT, echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
# Extract coordinates
X <- proc$X       # SMP (fitted)
Y <- proc$Yrot    # SNP (rotated)

# Plot empty frame with desired axis labels
plot(
  X[,1], X[,2],
  type = "n",                 # don't plot points yet
  xlab = "Dimension 1",
  ylab = "Dimension 2",
  axes = FALSE,               # remove axes
  frame.plot = TRUE           # draws box around plot
)

# Add points
points(X[,1], X[,2], pch = 16, col = "blue")   # SMP
points(Y[,1], Y[,2], pch = 17, col = "red")    # SNP

# Add arrows connecting SMP → SNP
arrows(X[,1], X[,2], Y[,1], Y[,2], length = 0.1, col = "gray")

# Add legend
legend(
  "topright",
  legend = c("SMP", "SNP"),
  pch = c(16,17),
  col = c("blue","red")
)

# Add Procrustes correlation and p-value
corr_text <- paste0("Procrustes DAPC correlation = ", round(proc_test[["t0"]], 3),
                    ", p = ", proc_test[["signif"]])
mtext(corr_text, side = 3, line = 0.5, adj = 0)  # top-left corner

```

```{r CHH PROCRUSTES PCA ANALYSIS, include=FALSE}
# Select only numeric PC axes
pca_axes <- pca_scores[, grep("^PC", colnames(pca_scores))]
pca_snp_axes <- pca_snp_scores[, grep("^PC", colnames(pca_snp_scores))]

# Make sure row order is identical
all(rownames(pca_axes) == rownames(pca_snp_axes))  # should be TRUE

# Run Procrustes
proc_pca <- procrustes(pca_axes, pca_snp_axes, scale = TRUE)

# Summary
summary(proc_pca)

# Statistical test
proc_pca_test <- protest(pca_axes, pca_snp_axes, permutations = 999)
proc_pca_test
```

```{r CHH PROCRUSTES PCA PLOT, echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
# Extract coordinates
X <- proc_pca$X       # SMP (fitted)
Y <- proc_pca$Yrot    # SNP (rotated)

# Plot empty frame with desired axis labels
plot(
  X[,1], X[,2],
  type = "n",                 # don't plot points yet
  xlab = "Dimension 1",
  ylab = "Dimension 2",
  axes = FALSE,               # remove axes
  frame.plot = TRUE           # draws box around plot
)

# Add points
points(X[,1], X[,2], pch = 16, col = "blue")   # SMP
points(Y[,1], Y[,2], pch = 17, col = "red")    # SNP

# Add arrows connecting SMP → SNP
arrows(X[,1], X[,2], Y[,1], Y[,2], length = 0.1, col = "gray")

# Add legend
legend(
  "topright",
  legend = c("SMP", "SNP"),
  pch = c(16,17),
  col = c("blue","red")
)

# Add Procrustes correlation and p-value
corr_text <- paste0("Procrustes PCA correlation = ", round(proc_pca_test[["t0"]], 3),
                    ", p = ", proc_pca_test[["signif"]])
mtext(corr_text, side = 3, line = 0.5, adj = 0)  # top-left corner

```


## Identifying important CHH sites

Utilizando un Loading plot podemos visualizar cuáles son las posiciones CHH más importantes en la separación entre grupos en el eje LD1.

```{r CHH LOADING PLOT, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Which CHHs contribute most to separation?
loadings <- loadingplot(dapc_res$var.contr, axis = 1, thres = 0.003, lab.jitter = 1)
```



```{r CHH DIFFERENTIAL METHYLATION ANALYSIS, include=FALSE}
# I will create a function to then compare all against all

# Function for pairwise comparison
meth_diff <- 10
p_value <- 0.05
pairwise_diff <- function(meth_obj, g1, g2, min_diff = meth_diff, q_cutoff = p_value) {
  
  # Subset the object to only keep the two groups of interest
  sub_obj <- reorganize(
    meth_obj,
    sample.ids = getSampleID(meth_obj)[meth_obj@treatment %in% c(g1, g2)],
    treatment = meth_obj@treatment[meth_obj@treatment %in% c(g1, g2)]
  )
  
  # Run differential methylation
  diff <- calculateDiffMeth(
    sub_obj,
    overdispersion = "MN",
    test = "Chisq"
  )
  
  # Extract significant CHHs
  sig <- getMethylDiff(diff, difference = min_diff, qvalue = q_cutoff)
  
  return(list(all = diff, significant = sig))
}


# MAKE COMPARISONS

# GRAZED VS. UNGRAZED (WITHIN UNCLIPPED)
res_GUvsUU <- pairwise_diff(CHH_meth, g1 = 2, g2 = 3)
# UNCLIPPED VS. CLIPPED (WITHIN GRAZED)
res_GUvsGC <- pairwise_diff(CHH_meth, g1 = 2, g2 = 0)
 # UNCLIPPED VS. CLIPPED (WITHIN UNGRAZED)
 res_UUvsUC <- pairwise_diff(CHH_meth, g1 = 3, g2 = 1)
```


## Differential methylation analysis



Con la función "pairwise_diff()" de methylKit, podemos comparar el estado de metilación de cada citosina entre la población de cada tratamiento y la población control (Grazed es el control de Ungrazed y Unclipped es el control de Clipped). Obtenemos hipo- e hiper- metilaciones con respecto al control, y cada una de estas diferencias está asociada a un p-valor. En este caso hemos considerado como Differentially Methylated Positions (**DMPs**) cuando la **diferencia de metilación \> `r meth_diff`%** y el **p-valor \< `r p_value`**


```{r CHH exporting table of DMPs, include=FALSE}

# Apply function to all your result objects
DMPs_CHH_GUvsUU <- res_GUvsUU$significant
DMPs_CHH_GUvsGC <- res_GUvsGC$significant
DMPs_CHH_UUvsUC <- res_UUvsUC$significant


# Export results as tab-delimited text files
write.table(DMPs_CHH_GUvsUU, "DMPs_CHH_GUvsUU.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(DMPs_CHH_GUvsGC, "DMPs_CHH_GUvsGC.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(DMPs_CHH_UUvsUC, "DMPs_CHH_UUvsUC.txt", sep = "\t", quote = FALSE, row.names = FALSE)

```


```{r CHH VOLCANO PLOTS FOR VISUALIZING DMPs, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Volcano plot function
plot_volcano <- function(diff_obj, min_diff = 10, q_cutoff = 0.05, title = "") {
  df <- as.data.frame(diff_obj)
  
  # Calculate variables
  df$logq <- -log10(df$qvalue)
  df$status <- ifelse(abs(df$meth.diff) >= min_diff & df$qvalue <= q_cutoff, "sig", "ns")
  
  # Plot
  p <- ggplot(df, aes(x = meth.diff, y = logq)) +
    geom_point(aes(color = status), alpha = 0.6, size = 1.5, show.legend = FALSE) +
    scale_color_manual(values = c("sig" = "red", "ns" = "black")) +
    geom_vline(xintercept = c(-min_diff, min_diff), linetype = "dashed", color = "blue") +
    geom_hline(yintercept = -log10(q_cutoff), linetype = "dashed", color = "blue") +
    labs(
      x = "% Methylation Difference",
      y = expression(-log[10]("q-value")),
      title = title
    ) +
    theme_bw() +
    theme(
      axis.title = element_blank(),  # remove axis labels
      axis.text = element_text(size = 9),
      plot.title = element_text(size = 11, hjust = 0.5)
    )
  
  return(p)
}

# Generate volcano plots
p1 <- plot_volcano(res_GUvsUU$all, title = "Ungrazed (Unclipped)")
p2 <- plot_volcano(res_GUvsGC$all, title = "Clipping (Grazed)")
p3 <- plot_volcano(res_UUvsUC$all, title = "Clipping (Ungrazed)")


grid.arrange(
  p1, p2, p3, ncol = 3,
  left   = textGrob(expression(-log[10](q-value)), rot = 90, gp = gpar(fontsize = 12)),
  bottom = textGrob("% Methylation Difference", gp = gpar(fontsize = 12))
)

```

```{r CHH DMPs table, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Extract the significant results
sigUngrazed <- res_GUvsUU$significant
sigClipped <- res_GUvsGC$significant
sigUngrazedClipped <- res_UUvsUC$significant

# Count positives and negatives
n_posUngrazed <- sum(sigUngrazed$meth.diff > 0)
n_negUngrazed <- sum(sigUngrazed$meth.diff < 0)

n_posClipped <- sum(sigClipped$meth.diff > 0)
n_negClipped <- sum(sigClipped$meth.diff < 0)

n_posUngrazedClipped <- sum(sigUngrazedClipped$meth.diff > 0)
n_negUngrazedClipped <- sum(sigUngrazedClipped$meth.diff < 0)


# Build the table
meth_table <- data.frame(
  Condition = c("Hypermethylated", "Hypomethylated"),
  UngrazedU = c(n_posUngrazed, n_negUngrazed),
  ClippingG = c(n_posClipped, n_negClipped),
  ClippingU = c(n_posUngrazedClipped, n_negUngrazedClipped)
)

# Show nice table in R Markdown
knitr::kable(meth_table, align = "c", caption = "Differentially methylated sites")

```
