---
title: "P.lagopus dbRDA"
author: "Andrés Delgado Galán"
date: "2026-02-24"
output: html_document
---

```{r LOADING DATA AND PACKAGES, include=FALSE}
# --- Core packages ---
library(here)        # Reproducibilidad de rutas
library(tidyverse)   # Incluye ggplot2, dplyr, tidyr, readr, purrr, stringr, forcats, tibble
library(cluster)     # Matriz de disimilaridad

# --- Estadística ---
library(vegan)       # dbRDA (capscale), permanova (adonis) y Rsq
# install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis") # Para esto hay que tener la versión correcta de Rtools, cli, y devtools (mirar en la documentación)
#library(pairwiseAdonis)  # adonis por pares


# --- Visualización y clustering ---


# --- Genética ---
library(vcfR)        # read and manipulate VCF files
library(adegenet)

# Cargamos SMPs de los tres contextos
# avena_CpG_obj <- readRDS(here("avena_ncbi", "avena_ncbi_CpG_meth_matrix.rds"))
# avena_CHG_obj <- readRDS(here("avena_ncbi", "avena_ncbi_CHG_meth_matrix.rds"))
# avena_CHH_obj <- readRDS(here("avena_ncbi", "avena_ncbi_CHH_meth_matrix.rds"))

# avena_all_CTXT_obj <- rbind(avena_CpG_obj, avena_CHG_obj, avena_CHH_obj)

plantago_CpG_obj <- readRDS(here("plantago_pseudo", "plantago_pseudo_CpG_meth_matrix.rds"))
plantago_CHG_obj <- readRDS(here("plantago_pseudo", "plantago_pseudo_CHG_meth_matrix.rds"))
plantago_CHH_obj <- readRDS(here("plantago_pseudo", "plantago_pseudo_CHH_meth_matrix.rds"))

plantago_all_CTXT_obj <- rbind(plantago_CpG_obj, plantago_CHG_obj, plantago_CHH_obj)

# Cargamos la tabla de tratamientos
treatment_table <- read.table(file = here("plantago_treatment.csv"), header = T, sep = ";")
```

```{r CpG distance MATRIX, include=FALSE}
#Transpose methylation matrix (individuals have to be rows)
CpG_meth_matrix <- t(plantago_CpG_obj)

# Create distance matrix
CpG_dist_matrix <- 
      daisy(CpG_meth_matrix, metric = c("euclidean"),
      stand = FALSE) # si queremos estandarizar habría que usar TRUE
D <- as.matrix(CpG_dist_matrix)
heatmap(D, symm = TRUE)
hc <- hclust(CpG_dist_matrix)
plot(hc)
heatmap(D, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc), symm = TRUE)

# Nos aseguramos de que el orden de las muestras es el mismo que en la matriz de metilación
treatment <- treatment_table[match(attr(CpG_dist_matrix, "Labels"), treatment_table$id), ]

```

```{r CHG distance MATRIX, include=FALSE}
#Transpose methylation matrix (individuals have to be rows)
CHG_meth_matrix <- t(plantago_CHG_obj)

# Create distance matrix
CHG_dist_matrix <- 
      daisy(CHG_meth_matrix, metric = c("euclidean"),
      stand = FALSE) # si queremos estandarizar habría que usar TRUE
D <- as.matrix(CHG_dist_matrix)
heatmap(D, symm = TRUE)
hc <- hclust(CHG_dist_matrix)
plot(hc)
heatmap(D, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc), symm = TRUE)

# Nos aseguramos de que el orden de las muestras es el mismo que en la matriz de metilación
treatment <- treatment_table[match(attr(CHG_dist_matrix, "Labels"), treatment_table$id), ]

```

```{r CHH distance MATRIX, include=FALSE}
#Transpose methylation matrix (individuals have to be rows)
CHH_meth_matrix <- t(plantago_CHH_obj)

# Create distance matrix
CHH_dist_matrix <- 
      daisy(CHH_meth_matrix, metric = c("euclidean"),
      stand = FALSE) # si queremos estandarizar habría que usar TRUE
D <- as.matrix(CHH_dist_matrix)
heatmap(D, symm = TRUE)
hc <- hclust(CHH_dist_matrix)
plot(hc)
heatmap(D, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc), symm = TRUE)

# Nos aseguramos de que el orden de las muestras es el mismo que en la matriz de metilación
treatment <- treatment_table[match(attr(CHH_dist_matrix, "Labels"), treatment_table$id), ]

```

```{r all_CTXT distance MATRIX, include=FALSE}
#Transpose methylation matrix (individuals have to be rows)
all_CTXT_meth_matrix <- t(plantago_all_CTXT_obj)

# Create distance matrix
all_CTXT_dist_matrix <- 
      daisy(all_CTXT_meth_matrix, metric = c("euclidean"),
      stand = FALSE) # si queremos estandarizar habría que usar TRUE
D <- as.matrix(all_CTXT_dist_matrix)
heatmap(D, symm = TRUE)
hc <- hclust(all_CTXT_dist_matrix)
plot(hc)
heatmap(D, Rowv = as.dendrogram(hc), Colv = as.dendrogram(hc), symm = TRUE)

# Nos aseguramos de que el orden de las muestras es el mismo que en la matriz de metilación
treatment <- treatment_table[match(attr(all_CTXT_dist_matrix, "Labels"), treatment_table$id), ]

```

```{r prepare genetic data to include in constrained dbRDA, include=FALSE}
# read VCF
vcf <- read.vcfR(here("plantago_pseudo", "filtered_variants_cov5-1000_biallelic_MAF0.01_80pct.vcf.gz"))

# Convert to genlight (handles large SNP sets efficiently)
gl <- vcfR2genlight(vcf)    # converts to genlight; ploidy defaults to 2
indNames(gl) <- sub("_R1_trimmed_bismark_bt2_pe\\.sorted$", "", indNames(gl))


# compute allele dosage matrix
dos <- as.matrix(gl)  # genlight -> numeric matrix (0/1/2), NAs remain
# Remove some monomorphic positions
keep <- apply(dos, 2, function(x) {
    u <- unique(na.omit(x))
    length(u) > 1
})
sum(!keep)    # should show 79
dos_filtered <- dos[, keep]

#missing data handling / imputation (replace NA by locus median)
# impute by median allele dosage per SNP:
for(j in seq_len(ncol(dos_filtered))){
  nas <- is.na(dos_filtered[,j])
  if(any(nas)) dos_filtered[nas,j] <- median(dos_filtered[,j], na.rm=TRUE)
}

# PRINCIPAL COMPONENT ANALYSIS
pca <- prcomp(dos_filtered)

# cALCULA EL % QUE EXPLICA CADA PC
var_exp <- pca$sdev^2
perc_var <- var_exp / sum(var_exp) * 100
perc_var

# CREA UNA TABLA PARA VISUALIZAR QUÉ PCS SON IMPORTANTES
var_table <- data.frame(
  PC = paste0("PC", seq_along(perc_var)),
  Variance = var_exp,
  Percent = perc_var,
  Cumulative = cumsum(perc_var)
)

# Tabla con el porcentaje de variabilidad genética, total y acumulado, que explica cada PC
var_table

# Decidir con cuántos PCs nos quedamos:
cum_var <- cumsum(pca$sdev^2 / sum(pca$sdev^2) * 100)
which(cum_var <= 35)


# sELECCIONA LOS PCs QUE VAYAS A USAR
gen_pcs <- as.data.frame(pca$x[, 1:1])
colnames(gen_pcs) <- paste0("PC", 1:1)
var_explained <- var_table$Cumulative[var_table$PC == "PC1"]
var_txt <- sprintf("%.2f", var_explained)


# VER SI COINCIDE EL ORDEN DE LAS MUESTRAS EN EL PCA Y EN LOS METADATOS
all(rownames(gen_pcs) == treatment$id)

# COMBINAR PCs CON METADATOS
treatment_pcs <- cbind(treatment, gen_pcs)

# dbRDA
mod_pc <- capscale(
  CpG_dist_matrix ~ maternal * simulation + Condition(PC1:PC1),
  data = treatment_pcs,
  add = FALSE
)

# Significancia global
anova(mod_pc, permutations = 9999)

# Significancia por término (esto lo hacen en García-Seoane cuando el mod_pcelo global sale significativo)
anova(mod_pc, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby mod_pceling the effect of each predictor while accounting for the others using partial constrained dbRDA"
RsquareAdj(mod_pc)

# FDR
```

In these analyses we evaluate how much epigenetic variability is explained by the herbivory treatments and their interaction. We analysed overall epigenetic variability and separately for each context. We used the methylation percent of each SMP (`r nrow(plantago_CpG_obj) ` in CpG, `r nrow(plantago_CHH_obj) ` in CHH, `r nrow(plantago_CHH_obj) ` in CHH) to construct an euclidean epigenetic distance matrix between samples (`r ncol(plantago_CpG_obj) `) using cluster::daisy.

```{r CpG dbRDA CpG_mod1, include=FALSE}
# dbRDA
CpG_mod1 <- capscale(
  CpG_dist_matrix ~ maternal,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_CpG_mod1 <- anova(CpG_mod1, permutations = 9999)
Sig_CpG_mod1

# Significancia por término (esto lo hacen en García-Seoane cuando el CpG_mod1elo global sale significativo)
anova(CpG_mod1, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CpG_mod1eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CpG_mod1 <- RsquareAdj(CpG_mod1)
Rsq_CpG_mod1

# FDR


```

```{r CpG dbRDA CpG_mod2, include=FALSE}
# dbRDA
CpG_mod2 <- capscale(
  CpG_dist_matrix ~ maternal * simulation,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_CpG_mod2 <- anova(CpG_mod2, permutations = 9999)
Sig_CpG_mod2

# Significancia por término (esto lo hacen en García-Seoane cuando el CpG_mod2elo global sale significativo)
anova(CpG_mod2, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CpG_mod2eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CpG_mod2 <- RsquareAdj(CpG_mod2)
Rsq_CpG_mod2

# FDR


```

```{r CpG dbRDA CpG_mod3, include=FALSE}
# dbRDA
CpG_mod3 <- capscale(
  CpG_dist_matrix ~ maternal * simulation + Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

# Significancia global
Sig_CpG_mod3 <- anova(CpG_mod3, permutations = 9999)
Sig_CpG_mod3

# Significancia por término (esto lo hacen en García-Seoane cuando el CpG_mod3elo global sale significativo)
anova(CpG_mod3, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CpG_mod3eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CpG_mod3 <- RsquareAdj(CpG_mod3)
Rsq_CpG_mod3

# FDR


```

```{r CpG dbRDA CpG_mod4, include=FALSE}

#  !!!
#  !!!

# MIRAR POR QUÉ EN MOUNGER (2022) USAN CONDITION... NO LE VEO EL SENTIDO PERO SEGURAMENTE LO TENGA

#  !!!
#  !!!

# dbRDA
CpG_mod4 <- capscale(
  CpG_dist_matrix ~ PC1, # Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

n_permut <- 9999
# Significancia global
Sig_CpG_mod4 <- anova(CpG_mod4, permutations = n_permut)
Sig_CpG_mod4

# Significancia por término (esto lo hacen en García-Seoane cuando el CpG_mod4elo global sale significativo)
anova(CpG_mod4, by = "term", permutations = n_permut)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CpG_mod4eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CpG_mod4 <- RsquareAdj(CpG_mod4)
Rsq_CpG_mod4

# FDR


```

```{r CHG dbRDA CHG_mod1, include=FALSE}
# dbRDA
CHG_mod1 <- capscale(
  CHG_dist_matrix ~ maternal,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_CHG_mod1 <- anova(CHG_mod1, permutations = 9999)
Sig_CHG_mod1

# Significancia por término (esto lo hacen en García-Seoane cuando el CHG_mod1elo global sale significativo)
anova(CHG_mod1, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHG_mod1eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHG_mod1 <- RsquareAdj(CHG_mod1)
Rsq_CHG_mod1

# FDR


```

```{r CHG dbRDA CHG_mod2, include=FALSE}
# dbRDA
CHG_mod2 <- capscale(
  CHG_dist_matrix ~ maternal * simulation,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_CHG_mod2 <- anova(CHG_mod2, permutations = 9999)
Sig_CHG_mod2

# Significancia por término (esto lo hacen en García-Seoane cuando el CHG_mod2elo global sale significativo)
anova(CHG_mod2, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHG_mod2eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHG_mod2 <- RsquareAdj(CHG_mod2)
Rsq_CHG_mod2

# FDR


```

```{r CHG dbRDA CHG_mod3, include=FALSE}
# dbRDA
CHG_mod3 <- capscale(
  CHG_dist_matrix ~ maternal * simulation + Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

# Significancia global
Sig_CHG_mod3 <- anova(CHG_mod3, permutations = 9999)
Sig_CHG_mod3

# Significancia por término (esto lo hacen en García-Seoane cuando el CHG_mod3elo global sale significativo)
anova(CHG_mod3, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHG_mod3eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHG_mod3 <- RsquareAdj(CHG_mod3)
Rsq_CHG_mod3

# FDR


```

```{r CHG dbRDA CHG_mod4, include=FALSE}

#  !!!
#  !!!

# MIRAR POR QUÉ EN MOUNGER (2022) USAN CONDITION... NO LE VEO EL SENTIDO PERO SEGURAMENTE LO TENGA

#  !!!
#  !!!

# dbRDA
CHG_mod4 <- capscale(
  CHG_dist_matrix ~ PC1, # Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

n_permut <- 9999
# Significancia global
Sig_CHG_mod4 <- anova(CHG_mod4, permutations = n_permut)
Sig_CHG_mod4

# Significancia por término (esto lo hacen en García-Seoane cuando el CHG_mod4elo global sale significativo)
anova(CHG_mod4, by = "term", permutations = n_permut)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHG_mod4eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHG_mod4 <- RsquareAdj(CHG_mod4)
Rsq_CHG_mod4

# FDR


```

```{r CHH dbRDA CHH_mod1, include=FALSE}
# dbRDA
CHH_mod1 <- capscale(
  CHH_dist_matrix ~ maternal,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_CHH_mod1 <- anova(CHH_mod1, permutations = 9999)
Sig_CHH_mod1

# Significancia por término (esto lo hacen en García-Seoane cuando el CHH_mod1elo global sale significativo)
anova(CHH_mod1, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHH_mod1eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHH_mod1 <- RsquareAdj(CHH_mod1)
Rsq_CHH_mod1

# FDR


```

```{r CHH dbRDA CHH_mod2, include=FALSE}
# dbRDA
CHH_mod2 <- capscale(
  CHH_dist_matrix ~ maternal * simulation,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_CHH_mod2 <- anova(CHH_mod2, permutations = 9999)
Sig_CHH_mod2

# Significancia por término (esto lo hacen en García-Seoane cuando el CHH_mod2elo global sale significativo)
anova(CHH_mod2, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHH_mod2eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHH_mod2 <- RsquareAdj(CHH_mod2)
Rsq_CHH_mod2

# FDR


```

```{r CHH dbRDA CHH_mod3, include=FALSE}
# dbRDA
CHH_mod3 <- capscale(
  CHH_dist_matrix ~ maternal * simulation + Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

# Significancia global
Sig_CHH_mod3 <- anova(CHH_mod3, permutations = 9999)
Sig_CHH_mod3

# Significancia por término (esto lo hacen en García-Seoane cuando el CHH_mod3elo global sale significativo)
anova(CHH_mod3, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHH_mod3eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHH_mod3 <- RsquareAdj(CHH_mod3)
Rsq_CHH_mod3

# FDR


```

```{r CHH dbRDA CHH_mod4, include=FALSE}

#  !!!
#  !!!

# MIRAR POR QUÉ EN MOUNGER (2022) USAN CONDITION... NO LE VEO EL SENTIDO PERO SEGURAMENTE LO TENGA

#  !!!
#  !!!

# dbRDA
CHH_mod4 <- capscale(
  CHH_dist_matrix ~ PC1, # Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

n_permut <- 9999
# Significancia global
Sig_CHH_mod4 <- anova(CHH_mod4, permutations = n_permut)
Sig_CHH_mod4

# Significancia por término (esto lo hacen en García-Seoane cuando el CHH_mod4elo global sale significativo)
anova(CHH_mod4, by = "term", permutations = n_permut)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby CHH_mod4eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_CHH_mod4 <- RsquareAdj(CHH_mod4)
Rsq_CHH_mod4

# FDR


```

```{r all_CTXT dbRDA all_CTXT_mod1, include=FALSE}
# dbRDA
all_CTXT_mod1 <- capscale(
  all_CTXT_dist_matrix ~ maternal,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_all_CTXT_mod1 <- anova(all_CTXT_mod1, permutations = 9999)
Sig_all_CTXT_mod1

# Significancia por término (esto lo hacen en García-Seoane cuando el all_CTXT_mod1elo global sale significativo)
anova(all_CTXT_mod1, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby all_CTXT_mod1eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_all_CTXT_mod1 <- RsquareAdj(all_CTXT_mod1)
Rsq_all_CTXT_mod1

# FDR


```

```{r all_CTXT dbRDA all_CTXT_mod2, include=FALSE}
# dbRDA
all_CTXT_mod2 <- capscale(
  all_CTXT_dist_matrix ~ maternal * simulation,
  data = treatment,
  add = FALSE
)

# Significancia global
Sig_all_CTXT_mod2 <- anova(all_CTXT_mod2, permutations = 9999)
Sig_all_CTXT_mod2

# Significancia por término (esto lo hacen en García-Seoane cuando el all_CTXT_mod2elo global sale significativo)
anova(all_CTXT_mod2, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby all_CTXT_mod2eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_all_CTXT_mod2 <- RsquareAdj(all_CTXT_mod2)
Rsq_all_CTXT_mod2

# FDR


```

```{r all_CTXT dbRDA all_CTXT_mod3, include=FALSE}
# dbRDA
all_CTXT_mod3 <- capscale(
  all_CTXT_dist_matrix ~ maternal * simulation + Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

# Significancia global
Sig_all_CTXT_mod3 <- anova(all_CTXT_mod3, permutations = 9999)
Sig_all_CTXT_mod3

# Significancia por término (esto lo hacen en García-Seoane cuando el all_CTXT_mod3elo global sale significativo)
anova(all_CTXT_mod3, by = "term", permutations = 9999)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby all_CTXT_mod3eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_all_CTXT_mod3 <- RsquareAdj(all_CTXT_mod3)
Rsq_all_CTXT_mod3

# FDR


```

```{r all_CTXT dbRDA all_CTXT_mod4, include=FALSE}

#  !!!
#  !!!

# MIRAR POR QUÉ EN MOUNGER (2022) USAN CONDITION... NO LE VEO EL SENTIDO PERO SEGURAMENTE LO TENGA

#  !!!
#  !!!

# dbRDA
all_CTXT_mod4 <- capscale(
  all_CTXT_dist_matrix ~ PC1, # Condition(PC1),
  data = treatment_pcs,
  add = FALSE
)

n_permut <- 9999
# Significancia global
Sig_all_CTXT_mod4 <- anova(all_CTXT_mod4, permutations = n_permut)
Sig_all_CTXT_mod4

# Significancia por término (esto lo hacen en García-Seoane cuando el all_CTXT_mod4elo global sale significativo)
anova(all_CTXT_mod4, by = "term", permutations = n_permut)

# R² ajustado (En García-Seoane la reportan en el texto pero no en la tabla)
# Boquete - "We also calculated individual adj. R2 for each predictorby all_CTXT_mod4eling the effect of each predictor while accounting for the others using partial constrained dbRDA"
Rsq_all_CTXT_mod4 <- RsquareAdj(all_CTXT_mod4)
Rsq_all_CTXT_mod4

# FDR


```

We conducted dbRDA using vegan::capscale with the epigenetic distance matrix as response and the herbivory treatments as predictors. To account for genetic variability, we summarized the SNP data (`r ncol(dos_filtered) ` SNPs) into principal components using PCA, and included the first PC (that explains `r var_txt`% of the genetic variability) as a condition in a constrained dbRDA. We tested the significance of the model using a permutation test with `r n_permut` permutations and obtained adjusted R2 values using the function RsquareAdj from the vegan package. We visualized the structure of the data in a 2-D space using the constrained ordination axes (RDA axes). We ran the following models for all contexts together and separately for each sequence context:

1.  Epigenetic distance \~ Maternal herbivory

2.  Epigenetic distance \~ Maternal herbivory \* Real-time herbivory

3.  Epigenetic distance \~ Maternal herbivory \* Real-time herbivory + Condition (PC)

4.  Epigenetic distance \~ PC


```{r Tabla unificada, echo=FALSE, results='hold', warning=FALSE, message=FALSE}
# Creamos una función para extraer los resultados de los test estadísticos para los 4 modelos
get_model_stats <- function(anova_obj, rsq_obj, model_label) {
  
  tab <- as.data.frame(anova_obj)
  
  data.frame(
    Model       = model_label,
    df          = tab$Df[1],
    SumOfSqs    = tab$Variance[1],
    F           = tab$F[1],
    p_value     = tab$`Pr(>F)`[1],
    adj_R2      = rsq_obj$adj.r.squared,
    stringsAsFactors = FALSE
  )
}

# Creamos una lista para almacenar los resultados de cada contexto
contextos <- c("CpG", "CHG", "CHH", "all_CTXT")
results_list <- list()

for(ctx in contextos) {
  # Construimos los nombres de los objetos dinámicamente
  sig_mod1 <- get(paste0("Sig_", ctx, "_mod1"))
  sig_mod2 <- get(paste0("Sig_", ctx, "_mod2"))
  sig_mod3 <- get(paste0("Sig_", ctx, "_mod3"))
  sig_mod4 <- get(paste0("Sig_", ctx, "_mod4"))
  
  rsq_mod1 <- get(paste0("Rsq_", ctx, "_mod1"))
  rsq_mod2 <- get(paste0("Rsq_", ctx, "_mod2"))
  rsq_mod3 <- get(paste0("Rsq_", ctx, "_mod3"))
  rsq_mod4 <- get(paste0("Rsq_", ctx, "_mod4"))
  
  # Creamos la tabla para este contexto
  tbl_ctx <- rbind(
    get_model_stats(sig_mod1, rsq_mod1, "ED ~ M"),
    get_model_stats(sig_mod2, rsq_mod2, "ED ~ M * S"),
    get_model_stats(sig_mod3, rsq_mod3, "ED ~ M * S + Condition(PC)"),
    get_model_stats(sig_mod4, rsq_mod4, "ED ~ PC")
  )
  
  # Añadimos FDR (calculado dentro de cada contexto)
  tbl_ctx$FDR <- p.adjust(tbl_ctx$p_value, method = "fdr")
  
  # Reordenamos columnas
  tbl_ctx <- tbl_ctx[, c("Model", "df", "SumOfSqs", "F", "p_value", "FDR", "adj_R2")]
  
  # Guardamos en la lista
  results_list[[ctx]] <- tbl_ctx
}

# Combinamos todos los contextos (ahora en el orden correcto)
results_tbl <- do.call(rbind, results_list)

footnote_text <- paste0(
  "The models include with and without adjusting for the variance explained by the genetic data (PC). PC, first principal component from genetic data (explaining " , var_txt, "% of the variation in scaled allele frequencies); df, degrees of freedom; F, value of the test statistic; FDR, false discovery rate; adj. R2, percent of variance explained, adjusted for multiple comparisons."
)

# Formatear p_value y FDR para mostrar <0.001 cuando sea necesario
results_tbl$p_value <- ifelse(results_tbl$p_value < 0.001, "<0.001", sprintf("%.3f", results_tbl$p_value))
results_tbl$FDR <- ifelse(results_tbl$FDR < 0.001, "<0.001", sprintf("%.3f", results_tbl$FDR))

# Tabla unificada con agrupaciones correctas
knitr::kable(
  results_tbl,
  digits = 3,
  format.args = list(scientific = FALSE),
  row.names = FALSE,
  caption = "Table 1. Results of the distance-based redundancy analysis (dbRDA) showing the percentage of epigenetic variance (ED) explained by maternal environment (M), real-time herbivory simulation (S), and/or their interaction (M:S) across different genomic contexts.",
  booktabs = TRUE) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::pack_rows("CpG", 1, 4) %>%
  kableExtra::pack_rows("CHG", 5, 8) %>%
  kableExtra::pack_rows("CHH", 9, 12) %>%
  kableExtra::pack_rows("All Contexts", 13, 16) %>%
  kableExtra::footnote(
    general = footnote_text,
    general_title = ""
  )
```



### CpG

```{r CpG RDA plots, echo=FALSE, results='hold', warning=FALSE, message=FALSE}

## --- Preparación común ---
treatment$group <- with(treatment,
  ifelse(maternal == "graz" & simulation == "clip", "0",
  ifelse(maternal == "ungraz" & simulation == "clip", "1",
  ifelse(maternal == "graz" & simulation == "unclip", "2", "3")))
)

treatment_colors <- c(
  "0" = "#FF69B4",
  "1" = "#00FF7F",
  "2" = "#8B3A62",
  "3" = "#008B45"
)

cols_alpha <- adjustcolor(
  treatment_colors[treatment$group],
  alpha.f = 0.6
)

## Función para dibujar cada RDA
plot_rda <- function(model, panel_label) {

  eig <- summary(model)$cont$importance[2, 1:2] * 100
  sc  <- scores(model, display = "sites", choices = 1:2)

  xlim <- range(sc[,1]) * 1.1
  ylim <- range(sc[,2]) * 1.1

  plot(
    sc,
    type = "p",
    pch = 19,
    col = cols_alpha,
    xlim = xlim,
    ylim = ylim,
    xlab = paste0("RDA1 (", round(eig[1], 1), "%)"),
    ylab = paste0("RDA2 (", round(eig[2], 1), "%)")
  )

  ## etiqueta del panel (A–D)
  mtext(panel_label, side = 3, adj = 0, line = 0.5, font = 2)
}

## --- Layout 2x2 ---
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot_rda(CpG_mod1, "A")
plot_rda(CpG_mod2, "B")
plot_rda(CpG_mod3, "C")
plot_rda(CpG_mod4, "D")


```

Fig. 2. Epigenetic differentiation plots. Ordination plots of the distance-based redundancy analysis (dbRDA) carried out to test for epigenetic differences in CG context with all four models: 1. ED \~ M (A), 2. ED \~ M \* S (B), 3. ED \~ M \* S + Condition(PC) (C), 4. ED \~ PC (D). RDA1: first constrained ordination axis; RDA2: second constrained ordination axis. Samples are colored by group (i.e.: maternal environment x real-time herbivory simulation).

### CHG

```{r CHG RDA plots, echo=FALSE, results='hold', warning=FALSE, message=FALSE}

## --- Preparación común ---
treatment$group <- with(treatment,
  ifelse(maternal == "graz" & simulation == "clip", "0",
  ifelse(maternal == "ungraz" & simulation == "clip", "1",
  ifelse(maternal == "graz" & simulation == "unclip", "2", "3")))
)

treatment_colors <- c(
  "0" = "#FF69B4",
  "1" = "#00FF7F",
  "2" = "#8B3A62",
  "3" = "#008B45"
)

cols_alpha <- adjustcolor(
  treatment_colors[treatment$group],
  alpha.f = 0.6
)

## Función para dibujar cada RDA
plot_rda <- function(model, panel_label) {

  eig <- summary(model)$cont$importance[2, 1:2] * 100
  sc  <- scores(model, display = "sites", choices = 1:2)

  xlim <- range(sc[,1]) * 1.1
  ylim <- range(sc[,2]) * 1.1

  plot(
    sc,
    type = "p",
    pch = 19,
    col = cols_alpha,
    xlim = xlim,
    ylim = ylim,
    xlab = paste0("RDA1 (", round(eig[1], 1), "%)"),
    ylab = paste0("RDA2 (", round(eig[2], 1), "%)")
  )

  ## etiqueta del panel (A–D)
  mtext(panel_label, side = 3, adj = 0, line = 0.5, font = 2)
}

## --- Layout 2x2 ---
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot_rda(CHG_mod1, "A")
plot_rda(CHG_mod2, "B")
plot_rda(CHG_mod3, "C")
plot_rda(CHG_mod4, "D")


```

Fig. 2. Epigenetic differentiation plots. Ordination plots of the distance-based redundancy analysis (dbRDA) carried out to test for epigenetic differences in CHG context with all four models: 1. ED \~ M (A), 2. ED \~ M \* S (B), 3. ED \~ M \* S + Condition(PC) (C), 4. ED \~ PC (D). RDA1: first constrained ordination axis; RDA2: second constrained ordination axis. Samples are colored by group (i.e.: maternal environment x real-time herbivory simulation).

### CHH

```{r CHH RDA plots, echo=FALSE, results='hold', warning=FALSE, message=FALSE}

## --- Preparación común ---
treatment$group <- with(treatment,
  ifelse(maternal == "graz" & simulation == "clip", "0",
  ifelse(maternal == "ungraz" & simulation == "clip", "1",
  ifelse(maternal == "graz" & simulation == "unclip", "2", "3")))
)

treatment_colors <- c(
  "0" = "#FF69B4",
  "1" = "#00FF7F",
  "2" = "#8B3A62",
  "3" = "#008B45"
)

cols_alpha <- adjustcolor(
  treatment_colors[treatment$group],
  alpha.f = 0.6
)

## Función para dibujar cada RDA
plot_rda <- function(model, panel_label) {

  eig <- summary(model)$cont$importance[2, 1:2] * 100
  sc  <- scores(model, display = "sites", choices = 1:2)

  xlim <- range(sc[,1]) * 1.1
  ylim <- range(sc[,2]) * 1.1

  plot(
    sc,
    type = "p",
    pch = 19,
    col = cols_alpha,
    xlim = xlim,
    ylim = ylim,
    xlab = paste0("RDA1 (", round(eig[1], 1), "%)"),
    ylab = paste0("RDA2 (", round(eig[2], 1), "%)")
  )

  ## etiqueta del panel (A–D)
  mtext(panel_label, side = 3, adj = 0, line = 0.5, font = 2)
}

## --- Layout 2x2 ---
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot_rda(CHH_mod1, "A")
plot_rda(CHH_mod2, "B")
plot_rda(CHH_mod3, "C")
plot_rda(CHH_mod4, "D")


```

Fig. 3. Epigenetic differentiation plots. Ordination plots of the distance-based redundancy analysis (dbRDA) carried out to test for epigenetic differences in CHH context with all four models: 1. ED \~ M (A), 2. ED \~ M \* S (B), 3. ED \~ M \* S + Condition(PC) (C), 4. ED \~ PC (D). RDA1: first constrained ordination axis; RDA2: second constrained ordination axis. Samples are colored by group (i.e.: maternal environment x real-time herbivory simulation).

### all_CTXT

```{r all_CTXT RDA plots, echo=FALSE, results='hold', warning=FALSE, message=FALSE}

## --- Preparación común ---
treatment$group <- with(treatment,
  ifelse(maternal == "graz" & simulation == "clip", "0",
  ifelse(maternal == "ungraz" & simulation == "clip", "1",
  ifelse(maternal == "graz" & simulation == "unclip", "2", "3")))
)

treatment_colors <- c(
  "0" = "#FF69B4",
  "1" = "#00FF7F",
  "2" = "#8B3A62",
  "3" = "#008B45"
)

cols_alpha <- adjustcolor(
  treatment_colors[treatment$group],
  alpha.f = 0.6
)

## Función para dibujar cada RDA
plot_rda <- function(model, panel_label) {

  eig <- summary(model)$cont$importance[2, 1:2] * 100
  sc  <- scores(model, display = "sites", choices = 1:2)

  xlim <- range(sc[,1]) * 1.1
  ylim <- range(sc[,2]) * 1.1

  plot(
    sc,
    type = "p",
    pch = 19,
    col = cols_alpha,
    xlim = xlim,
    ylim = ylim,
    xlab = paste0("RDA1 (", round(eig[1], 1), "%)"),
    ylab = paste0("RDA2 (", round(eig[2], 1), "%)")
  )

  ## etiqueta del panel (A–D)
  mtext(panel_label, side = 3, adj = 0, line = 0.5, font = 2)
}

## --- Layout 2x2 ---
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot_rda(all_CTXT_mod1, "A")
plot_rda(all_CTXT_mod2, "B")
plot_rda(all_CTXT_mod3, "C")
plot_rda(all_CTXT_mod4, "D")


```

Fig. 4. Epigenetic differentiation plots. Ordination plots of the distance-based redundancy analysis (dbRDA) carried out to test for epigenetic differences in all_CTXT context with all four models: 1. ED \~ M (A), 2. ED \~ M \* S (B), 3. ED \~ M \* S + Condition(PC) (C), 4. ED \~ PC (D). RDA1: first constrained ordination axis; RDA2: second constrained ordination axis. Samples are colored by group (i.e.: maternal environment x real-time herbivory simulation).

```{r CpG PERMANOVA, include=FALSE}
# PERMANOVA
vegan::adonis2(CpG_dist_matrix ~ maternal * simulation,
  data = treatment,
  permutations = 9999,
  method = "euclidean",
  by = "terms")   # Probar a correr con by = NULL para ver la significación global

```

